\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-21 1-intro.Rmd
notes of another round of changes to make in the introduction

Spatio-temporal data concerns variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube with three axes: spatial identifier, temporal identifier, and variable. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows four examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data layout: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.", fig.align='center', out.height = "40%"}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

@@

>>>>

Spatio-temporal data involves variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube with three axes: spatial identifier, temporal identifier, and variable. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows four examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data layout: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.", fig.align='center'}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

@@

\textcolor{red}{start with "we need something to do exploratory data analysis for spatio-temporal data", then why it is hard to work with. This section needs to be abstract a bit further.}

### 2022-02-21 main.Rmd
notes of another round of changes to make in the introduction

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolas@langrene.com
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Exploring spatio-temporal data can sometimes be discouraging given the various different formats spatio-temporal data can present themselves in practice. Moreover, the two observational units naturally exist in spatio-temporal data can cost analysts many time and efforts on organising variables and results and switching between a single table and multiple tables in the analysis. In this paper, a new data structure, \code{cubble}, is proposed for wrangling and visualising spatio-temporal data. The new data structure uses a single data object to organise spatial and temporal variables in two forms so that they can be wrangled separately while keep syncronised. Advanced considerations are given to illustrate data with hierarchical structure, data matching, how cubble fits with the interactive graphics pipeline, and spatio-temporal transformation. Examples are given to analysing Australia meteorology data, river level data, and climate reanalysis (ERA5). 
keywords:
  # at least one keyword must be supplied
  formatted: [spatio temporal,  "\\proglang{R}", climate weather station, data analysis]
  plain:     [spatio temporal, R, climate weather station, data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

>>>>

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolas@langrene.com
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >

  \textcolor{red}{start with describing the data: spatial and temporal, then what's the problem, then propose cubble.}

  \textcolor{red}{Exploring spatio-temporal data can sometimes be discouraging given the various different formats spatio-temporal data can present themselves in practice}. Moreover, the two observational units naturally exist in spatio-temporal data \textcolor{red}{can cost analysts many time and efforts} on organising variables and results and switching between a single table and multiple tables in the analysis. In this paper, a new data structure, \code{cubble}, is proposed for wrangling and visualising spatio-temporal data. The new data structure uses a single data object to organise spatial and temporal variables in two forms so that they can be wrangled separately while keep syncronised. Advanced considerations are given to illustrate data with hierarchical structure, data matching, how cubble fits with the interactive graphics pipeline, and spatio-temporal transformation. Examples are given to analysing Australia meteorology data, river level data, and climate reanalysis (ERA5). 
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

### 2022-02-21 main.tex
notes of another round of changes to make in the introduction

\Abstract{
Exploring spatio-temporal data can sometimes be discouraging given the
various different formats spatio-temporal data can present themselves in
practice. Moreover, the two observational units naturally exist in
spatio-temporal data can cost analysts many time and efforts on
organising variables and results and switching between a single table
and multiple tables in the analysis. In this paper, a new data
structure, \code{cubble}, is proposed for wrangling and visualising
spatio-temporal data. The new data structure uses a single data object
to organise spatial and temporal variables in two forms so that they can
be wrangled separately while keep syncronised. Advanced considerations
are given to illustrate data with hierarchical structure, data matching,
how cubble fits with the interactive graphics pipeline, and
spatio-temporal transformation. Examples are given to analysing
Australia meteorology data, river level data, and climate reanalysis
(ERA5).
}

\Keywords{spatio temporal, \proglang{R}, climate weather station, data
analysis}
\Plainkeywords{spatio temporal, R, climate weather station, data
analysis}

@@

Spatio-temporal data concerns variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time interval. This type of data can naturally
been considered as a data cube with three axes: spatial identifier,
temporal identifier, and variable. In this cubic formation, space only
occupies one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows four examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

{\centering \includegraphics[width=1\linewidth,height=0.4\textheight]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.001}

>>>>

\Abstract{
\textcolor{red}{start with describing the data: spatial and temporal, then what's the problem, then propose cubble.}
\textcolor{red}{Exploring spatio-temporal data can sometimes be discouraging given the various different formats spatio-temporal data can present themselves in practice}.
Moreover, the two observational units naturally exist in spatio-temporal
data \textcolor{red}{can cost analysts many time and efforts} on
organising variables and results and switching between a single table
and multiple tables in the analysis. In this paper, a new data
structure, \code{cubble}, is proposed for wrangling and visualising
spatio-temporal data. The new data structure uses a single data object
to organise spatial and temporal variables in two forms so that they can
be wrangled separately while keep syncronised. Advanced considerations
are given to illustrate data with hierarchical structure, data matching,
how cubble fits with the interactive graphics pipeline, and
spatio-temporal transformation. Examples are given to analysing
Australia meteorology data, river level data, and climate reanalysis
(ERA5).
}

\Keywords{spatial, temporal, spatio
temporal, \proglang{R}, environmental data, exploratory data analysis}
\Plainkeywords{spatial, temporal, spatio temporal, R, environmental
data, exploratory data analysis}

@@

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time interval. This type of data can naturally
been considered as a data cube with three axes: spatial identifier,
temporal identifier, and variable. In this cubic formation, space only
occupies one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows four examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.001}
